Search techniques for gaussian mixture modelling with Minimum Message Length, with applications to the origin of our Galaxy

### 2018-09-10 article/ms.tex
merge

While we can evaluate the objective function $I(\vec{\data}|\vec{\psi})$, 
where for the sake of brevity we denote $\vec{\psi} = \{K, N, D, \vec{\weight}, 
\vecmean, \veccov\}$, and we can meaningfully compare models with different 
numbers of components or parameter values, we require a search strategy in 
order to efficiently find the global solution: the minimum message length 
over a class of finite mixture models (where $N > K$). Before we introduce and 
compare search strategies, let us first make some comments and quantitative 
descriptions on how the message length $I$ will vary with the model parameters. 
Specifically, we will re-arrange Equation \ref{eq:objective_function} to be a 
sum of three smaller functions, and we will derive limits and predictions for 
those quantites. In an ideal scenario, those limits and predictions would be 
able to provide us with an estimate of the message length for a Gaussian 
mixture before computing the mixture (i.e., before performing 
expectation-maximization.We begin by re-casting Eq. \ref{eq:objective_function} 
as
\begin{eqnarray}
I(\vec{\data}|\vec{\psi}) = I_{1}(K, N, D, \vec{\weight}) 
                          + I_{2}(K, N, \veccov)
                          + I_{3}(\vec{\psi}) \quad ,
\end{eqnarray}
\noindent{}where:
\begin{eqnarray}
    I_{1} &=& \left(1-\frac{D}{2}\right)K\log{2} + \log{\Gamma(K)} + \frac{1}{4}\left[2\left(K - 1\right) + KD\left(D+3\right)\right]\log{N} \nonumber \\
          &+& \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right) + \left(\frac{D(D+3) - 2}{4}\right)\sum_{k=1}^{K}\log{\weight_k} \todo{+ \kappa} \quad , \label{eq:I1} \\
    I_{2} &=& -\left(1 + \frac{D}{2}\right)\sum_{k=1}^{K}\log{|\veccov_k|} \quad {\textrm and},\\
    I_{3} &=& -\log\likelihood\left(\vec{\data}|\vec{\theta}\right) - DN\log{\sigma_{\datum}} \quad .
\end{eqnarray}
The nomenclature for $I_1$, $I_2$ and $I_3$ bear no relation to the two-part 
message that we introduced earlier to describe message length. Rather we are
simply re-arranging the objective function because, as we will show, there are
useful theoretical bounds and predictions that we can make for each part of
this re-arranged expression.

@@

\begin{enumerate}
    \item The weights are distributed uniformly such that $w_k = 1/K$ and
          \begin{equation}
                \sum_{k=1}^{K}\log\weight_k = \sum_{k=1}^{K}\log{\frac{1}{K}} =-K\log{K} \quad .
          \end{equation}
    \item The weights are distributed such that $K - 1$ components each 
          describe just one data point $w_{1..K-1} = 1/N$, and the remaining 
          component describes all other data points $w_{K} = 1 - (K-1)/N$
          as $\sum_{k=1}^{K}\weight_k = 1$. At this extreme,
          \begin{equation}
                \sum_{k=1}^{K}\log\weight_k = (K-1)\log\frac{1}{N} + \log\left(1 - \frac{K - 1}{N}\right)
          \end{equation}
          \noindent{}and by making use of the identity 
            $\log{(a-b)} = \log{a} + \log{(1 - \frac{b}{a})}$
          we can simplify this expression to:
          \begin{equation}
                \sum_{k=1}^{K}\log\weight_k = \log{(N-K+1)} - K\log{N} \quad .
          \end{equation}
\end{enumerate}

\noindent{}Thus, for any finite mixture of $K$ Gaussian components we have 
derived the bounds
\begin{equation}
-K\log{K} \quad \leq \quad \sum_{k=1}^{K}\log\weight_k \quad \leq \quad - K\log{K} + \log(N + 1 - K)
\end{equation}
\noindent{}which gives us analytic expressions for the upper and lower value 
of $I_1$, regardless of the values of $\vec{\weight}$, for any mixture, and
bounds on $I_{1}$ can be derived from just $\{K, N, D\}$.

@@

There is no finite upper bound on $I_{3}$ because the negative log-likelihood
$-\log\likelihood$ will approach infinity for an increasingly poor fit to the
data. A useful lower limit, however, does exist. The negative log-likelihood
in Eq. \ref{eq:log-likelihood} can be expanded as,
\begin{equation}
    -\log\mathcal{L}\left(\vec{\data}|\vec{\theta}\right) = \sum_{n=1}^{N}\log\sum_{k=1}^{K}\frac{\weight_k}{\sqrt{(2\pi)^{D}|\vec{\cov}_k|}}\exp\left[-\frac{1}{2}\left(\vec{\data}_n - \vecmean_k\right)^\intercal{}\vec{\cov}_{k}^{-1}\left(\vec{\data}_n - \vecmean_k\right)\right]
\end{equation}
\noindent{}and we can define
\begin{equation}
\chi^{2}_{nk} = \left(\vec{\data}_{n} - \vecmean_{k}\right)^\intercal{}\vec{\cov}_{k}^{-1}\left(\vec{\data}_{n} - \vecmean_{k}\right) \quad .
\end{equation}
For the sake of placing a lower bound on $I_{3}$, let us assume that each data
point will be `mostly' explained by one of the components in the 
mixture\footnote{We model the data with partial assignment, but here we merely 
seek to place a lower bound on the negative log-likelihood.}, but we do not 
necessarily know which component, or what the parameters of that component are. 
In that situation, we can state that the component that \emph{does} explain
the data will have $\chi_{nk}^2 \sim \textrm{O}\left(D\right)$ and $\exp(-\frac{1}{2}\chi^2) \rightarrow 1$
whereas the other components that do not explain the data will have 
$\chi_{nk}^2 >> D$ such that $\exp(-\frac{1}{2}\chi^2) \rightarrow 0$. With
this simplification we can write an approximation for the negative log-likelihood
that does not require knowledge about which data points are associated to what
component, because the number of times that the expression $\exp(-\frac{1}{2}\chi^2)$
will be non-zero is for the $k$th mixture component will be $N\weight_k$. The
approximate lower bound is:
\begin{eqnarray}
    -\log\likelihood\left(\vec{\data}|\vec{\theta}\right) & \gtrsim & -\sum_{k=1}^{K}N\weight_{k}\log\left[\frac{\weight_k}{\sqrt{(2\pi)^{D}|\vec{\cov}_k|}}\exp\left(-\frac{1}{2}{D}\right)\right] \\
    -\log\likelihood\left(\vec{\data}|\vec{\theta}\right) & \gtrsim & -N\sum_{k=1}^{K}\weight_{k}\log{\weight_k} + \frac{1}{2}N\sum_{k=1}^{K}\weight_{k}\log{|\vec{\cov}_k|} + \frac{1}{2}ND\left(1 + \log{2\pi}\right) \quad . \nonumber
    \label{eq:logl-lower-bound}
\end{eqnarray}
\noindent{}Like before, we can consider the lower bound on $-N\sum_{k=1}^{K}\weight_{k}\log\weight_k$
for any mixture of $K$ components by setting the weight of $K - 1$ mixtures
to the smallest realistic values, where those mixtures encapsulate only two
data points and $\todo{w_{1} = w_2 = w_{\cdots} = w_{K-1} = 2/N}$:

\begin{eqnarray}
-N\sum_{k=1}^{K}\weight_{k}\log\weight_k &=& -N\left[(K-1)\left(\frac{1}{N}\log\frac{1}{N}\right) + \left(1 - \frac{K-1}{N}\right)\log\left(1 - \frac{K-1}{N}\right)\right] \nonumber \\
-N\sum_{k=1}^{K}\weight_{k}\log\weight_k &=& (K - 1)\log{N} + (K - 1 - N)\log\left(1-\frac{K-1}{N}\right) \nonumber \\
-N\sum_{k=1}^{K}\weight_{k}\log\weight_k &=& N\log{N} + (K - 1 - N)\log\left(N-K+1\right) \quad .
\end{eqnarray}

>>>>

While we can evaluate the objective function $I(\vec{\data}|\vec{\psi})$, 
where for the sake of brevity we denote $\vec{\psi} = \{K, N, D, \vec{\weight}, 
\vecmean, \veccov\}$, and we can meaningfully compare models with different 
numbers of components or parameter values, we require a search strategy in 
order to efficiently find the global solution: the minimum message length 
over a class of finite mixture models (where $N > K$). Before we introduce and 
compare search strategies, let us first make some comments and quantitative 
descriptions on how the message length $I$ will vary with the model parameters. 
Specifically, we will re-arrange Equation \ref{eq:objective_function} to be a 
sum of three smaller functions, and we will derive limits and predictions for 
those quantites. In an ideal scenario, those limits and predictions would be 
able to provide us with an estimate of the message length for a Gaussian 
mixture before computing the mixture (i.e., before performing 
expectation-maximization.We begin by re-casting Eq. \ref{eq:objective_function} 
as
\begin{eqnarray}
I(\vec{\data}|\vec{\psi}) = I_{1}(K, N, D, \vec{\weight}) 
                          + I_{2}(K, N, \veccov)
                          + I_{3}(\vec{\psi}) \quad ,
\end{eqnarray}
\noindent{}where:
\begin{eqnarray}
    I_{1} &=& \left(1-\frac{D}{2}\right)K\log{2} + \log{\Gamma(K)} + \frac{1}{4}\left[2\left(K - 1\right) + KD\left(D+3\right)\right]\log{N} \nonumber \\
          &+& \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right) + \left(\frac{D(D+3) - 2}{4}\right)\sum_{k=1}^{K}\log{\weight_k} \todo{+ \kappa} \quad , \label{eq:I1} \\
    I_{2} &=& -\left(1 + \frac{D}{2}\right)\sum_{k=1}^{K}\log{|\veccov_k|} \quad \textrm{and},\\
    I_{3} &=& -\log\likelihood\left(\vec{\data}|\vec{\theta}\right) - DN\log{\mathcal{E}} \quad .
\end{eqnarray}
The nomenclature for $I_1$, $I_2$ and $I_3$ bear no relation to the two-part 
message that we introduced earlier to describe message length. Rather we are
simply re-arranging the objective function because, as we will show, there are
useful theoretical bounds and predictions that we can make for each part of
this re-arranged expression.

@@

\begin{enumerate}
    \item The weights are distributed uniformly such that $w_k = 1/K$ and
          \begin{equation}
                \sum_{k=1}^{K}\log\weight_k = \sum_{k=1}^{K}\log{\frac{1}{K}} =-K\log{K} \quad .
          \end{equation}
    \item The weights are distributed such that $K - 1$ components each 
          describe just two data points $w_{1..K-1} = 2/N$, and the remaining 
          component describes all other data points $w_{K} = 1 - 2(K-1)/N$
          as $\sum_{k=1}^{K}\weight_k = 1$. At this extreme,
          \begin{equation}
                \sum_{k=1}^{K}\log\weight_k = 2(K-1)\log\frac{2}{N} 
                		    + \log\left(1 - \frac{2(K - 1)}{N}\right)
          \end{equation}
          \noindent{}and by making use of the identity 
            $\log{(a-b)} = \log{a} + \log{(1 - \frac{b}{a})}$:
          \begin{equation}
                \sum_{k=1}^{K}\log\weight_k = 2K\left(\log{2} - 1\right) - 2\log{2} + \log{N} + \log{\left(N - 2K + 2\right)} \quad .
          \end{equation}
\end{enumerate}

\noindent{}Thus, for any finite mixture of $K$ Gaussian components we have 
derived the bounds
\begin{eqnarray}
\sum_{k=1}^{K}\log\weight_k & \geq & -K\log{K} \nonumber \\
\sum_{k=1}^{K}\log\weight_k & \leq & 2K\left(\log{2} - 1\right) - 2\log{2} + \log{N} + \log{\left(N - 2K + 2\right)} \nonumber
\end{eqnarray}
\noindent{}which gives us analytic expressions for the upper and lower value 
of $I_1$, regardless of the values of $\vec{\weight}$, for any mixture, and
bounds on $I_{1}$ can be derived from just $\{K, N, D\}$.

@@

There is no finite upper bound on $I_{3}$ because the negative log-likelihood
$-\log\likelihood$ will approach infinity for an increasingly poor fit to the
data. A useful lower limit, however, does exist. The negative log-likelihood
in Eq. \ref{eq:log-likelihood} can be expanded as,
\begin{equation}
    -\log\mathcal{L}\left(\vec{\data}|\vec{\theta}\right) = \sum_{n=1}^{N}\log\sum_{k=1}^{K}\frac{\weight_k}{\sqrt{(2\pi)^{D}|\vec{\cov}_k|}}\exp\left[-\frac{1}{2}\left(\vec{\data}_n - \vecmean_k\right)^\intercal{}\vec{\cov}_{k}^{-1}\left(\vec{\data}_n - \vecmean_k\right)\right]
\end{equation}
\noindent{}and we can define
\begin{equation}
\chi^{2}_{nk} = \left(\vec{\data}_{n} - \vecmean_{k}\right)^\intercal{}\vec{\cov}_{k}^{-1}\left(\vec{\data}_{n} - \vecmean_{k}\right) \quad .
\end{equation}
For the sake of placing a lower bound on $I_{3}$, let us assume that each data
point will be `mostly' explained by one of the components in the 
mixture\footnote{We model the data with partial assignment, but here we merely 
seek to place a lower bound on the negative log-likelihood.}, but we do not 
necessarily know which component, or what the parameters of that component are. 
In that situation, we can state that the component that \emph{does} explain
the data will have $\chi_{nk}^2 \sim \textrm{O}\left(D\right)$ and $\exp(-\frac{1}{2}\chi^2) \rightarrow 1$
whereas the other components that do not explain the data will have 
$\chi_{nk}^2 >> D$ such that $\exp(-\frac{1}{2}\chi^2) \rightarrow 0$. With
this simplification we can write an approximation for the negative log-likelihood
that does not require knowledge about which data points are associated to what
component, because the number of times that the expression $\exp(-\frac{1}{2}\chi^2)$
will be non-zero is for the $k$th mixture component will be $N\weight_k$. The
approximate lower bound is:
\begin{eqnarray}
    -\log\likelihood\left(\vec{\data}|\vec{\theta}\right) & \gtrsim & -\sum_{k=1}^{K}N\weight_{k}\log\left[\frac{\weight_k}{\sqrt{(2\pi)^{D}|\vec{\cov}_k|}}\exp\left(-\frac{1}{2}{D}\right)\right] \\
    -\log\likelihood\left(\vec{\data}|\vec{\theta}\right) & \gtrsim & -N\sum_{k=1}^{K}\weight_{k}\log{\weight_k} + \frac{1}{2}N\sum_{k=1}^{K}\weight_{k}\log{|\vec{\cov}_k|} + \frac{1}{2}ND\left(1 + \log{2\pi}\right) \quad . \nonumber
    \label{eq:logl-lower-bound}
\end{eqnarray}
\noindent{}Like before, we can consider the lower bound on $-N\sum_{k=1}^{K}\weight_{k}\log\weight_k$
for any mixture of $K$ components by setting the weight of $K - 1$ mixtures
to the smallest realistic values, where those mixtures encapsulate only two
data points and $w_{1} = w_2 = w_{\cdots} = w_{K-1} = 2/N$ such that
%\begin{eqnarray}
%	\sum_{k=1}^{K}\weight_{k}\log\weight_k = \left[(K-1)\left(\frac{2}{N}\log{\frac{2}{N}}\right) + \left(1-\frac{2(K-1)}{N}\right)\log{\left(1-\frac{2(K-1)}{N}\right)}\right] \nonumber
%\end{eqnarray}`
%\noindent{}and
\begin{eqnarray}
	-N\sum_{k=1}^{K}\weight_{k}\log\weight_k \geq 2(1-K)\log{2} + N\log{N} - (N-2K+2)\log{\left(N - 2K + 2\right)}
\end{eqnarray}
%  log(a-b) = log(a) + log(1 - b/a)
% log(1 - b/a) = log(a-b) - log(a)
% log(1 - b/a) = log(N - 2K + 2) - log(N)

% 2(1-K)log(2) - 2(1-K)log(N) + (2K - N - 2)log(N - 2K + 2) - (2K - N - 2)log(N)
% 2(1-K)log(2) - 2log(N) + 2Klog(N) + (2K - N - 2)log(N - 2K + 2) - (2K - N - 2)log(N)
% 2(1-K)log(2) + (2K - 2)log(N) + (2K - N - 2)log(N - 2K + 2) - (2K - 2 - N)log(N)
% 2(1-K)log(2) + Nlog(N) + (2K - N - 2)log(N - 2K + 2)
